The Talking Heads experiment

### 2015-04-28 chap6/chap6.tex
fixed remaining widows and orphans
format relevant papers section

The speaker first plays a Discrimination Game traversing the 
semantic side of the square going from the referent 
{\itshape rectangle-1} to a possible meaning [\textsc{gray} 0.5–1.0]. 
He then plays a Naming Game traversing the lexical side of 
of the square to find the word `pokuneso' for this chosen meaning. 
The hearer traverses the lexical side of the triangle in 
the other direction to interpret the word `pokuneso' as
{}[\textsc{gray} 0.5–1.0], and then identifies the referent by 
filtering the objects in the context with this meaning. 
Only rectangle-1 remains, so the game succeeds. 
The whole game is reported by the commentator as follows: 
\begin{verbatim}
Game 500
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 3 objects: 
       rectangle-0 rectangle-1 rectangle-2
  a2 chooses rectangle-1 as the topic 
  a2 categorises the topic as [GRAY 0.5–1.0]
  a2 says: pokuneso
  a1 interprets pokuneso as [GRAY 0.5–1.0]
  a1 points to rectangle-1
  a2 signals OK 
\end{verbatim}
The game is perfectly successful because both 
agents associate the word `okuneso' with 
{}[\textsc{gray} 0.5–1.0] (dark) and they perceive the scene 
in the same way.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l }
\lsptoprule
{\itshape Meaning}&{\itshape Word}&{\itshape Translation} & {\bfshape  a1}&{\bfshape  a2} \\ \midrule
{}[\textsc{hpos} 0.0–0.5] & vapola&left&-&0.1\\ 
{}[\textsc{hpos} 0.5–1.0]& gonapa&right &0.1&-\\ 
{}[\textsc{height} 0.0–0.5]&suwaxugo&short &0.6&0.8\\ 
{}[\textsc{height} 0.5–1.0]& kusone&tall &0.4&0.5\\ 
{}[\textsc{width} 0.0–0.5]&bepupepa&narrow &0.1&0.1\\ 
{}[\textsc{width} 0.0–0.25]&kutaki&very narrow &-&0.1\\ 
{}[\textsc{width} 0.5–1.0]& zikorika&wide &0.0&0.3\\ 
{}[\textsc{gray} 0.0–0.5]& fesasado&light &0.5&0.7\\ 
{}[\textsc{gray} 0.5–1.0]& pokuneso&dark &0.8&0.9\\ 
{}[\textsc{area} 0.5–1.0]& mafanoda&large &0.1&0.1\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:lex100}. Complete lexicon of {\bfshape  a1} and {\bfshape  a2} after 100 games.}
\end{center}
\end{table}
We see that, at this point, the agents have lexicalised 
only the most general distinctions, such as `dark' (\emph{pokuneso}) 
versus `light' (\emph{fesasado}) or `short' (\emph{suwaxugo}) versus `tall' 
(\emph{kusone}). Words for the grayscale and height dimensions
have the strongest scores, although this is purely accidental. 
When we would start another simulation from scratch,
we would end up with different words and perhaps 
other distinctions would be more successful.

@@

We see that words for  basic distinctions have further established
themselves. Words for `short' (\emph{suwaxugo}) and `tall' (\emph{kusone}), or `light' (\emph{fesasado}) and `dark' (\emph{pokuneso}) now have scores of 1.0. Words for more refined categories, like `very short' (\emph{tawube}) or `very narrow' (\emph{kutaki}), are beginning to establish themselves.

@@

Note that this simulation is very different from the 
ones shown in the previous chapter. The agents now get 
only feedback through overt selection of the referent. 
The hearer points to the identified referent and the speaker 
decides on the outcome of the game based on this 
non-verbal information, but speaker and hearer do not 
know whether they have used the same meaning or not. 
Very often there are alternative ways
to conceptualise reality, so even if agents would have
completely shared ontologies, there is still the possibility 
of guessing the wrong meaning. I have called this
the gavagai-problem, inspired by the philosopher
Quine, who tells the story of the anthropologist 
puzzled by the word `gavagai' uttered by a native 
in an undecoded language. Does `gavagai' mean rabit, animal 
scurrying by, the direction in which I will go 
now, or white furry object? 
The child who is acquiring a lexicon has
exactly the same problem. It explains why 
overextensions or underextensions
are seen in a child's first words. For example, 
the word for orange is applied to any small circular round 
object, including a ball, or a doorknob.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l }
\lsptoprule
{\itshape Obj} & Hpos & Vpos & Height & Width & Gray & Area \\ \midrule
0 & 0.51 & 0.98 & 0.90 & 0.47 & 0.79 & 0.60\\ 
1 & 0.75 & 0.68 & 0.26 & 0.09 & 0.14 & 0.20\\ 
2 & 0.76 & 0.54 & 0.56 & 0.26 & 0.94 & 0.36\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game77}Sensory data for game 77 after sensor-scaling.}
\end{center}
\end{table}
The speaker is again {\bfshape  a2}. \textsc{height} is the most salient channel. 
{\bfshape  a2} has a word for [\textsc{height} 0.75–1.0] (very tall), namely
`bowaluro', and uses it. The hearer {\bfshape  a1} 
does not know the 
word, conceptualises the scene, and arrives at the 
same category, so the hearer stores the new word
with the same meaning as the speaker. 
\begin{verbatim}
Game 77
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 3 objects: 
       rectangle-0 rectangle-1 rectangle-2
  a2 chooses rectangle-1 as the topic 
  a2 categorises the topic as [HEIGHT 0.75–1.0]
  a2 says: `bowaluro'
  a1 does not know `bowaluro'
  a1 says: `bowaluro?'
  a2 points to rectangle-1
  a1 categorises the topic as [HEIGHT 0.75–1.0]
  a1 stores `bowaluro' as [HEIGHT 0.75–1.0]
\end{verbatim}
The reason why {\bfshape  a1} has guessed the right 
meaning of `bowaluro' is because both agents use
only the most salient channel and they both share
the same perception of reality. We will soon see 
that if these constraints are not valid, agents 
are not always so lucky and hence multiple meanings
start to circulate for the same word.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l }
\lsptoprule
{\itshape Obj}&Hpos&Vpos&Height&Width&Gray&Area \\ \midrule
0 &0.51 & 0.71 & 0.64 & 0.61 & 0.80 & 0.56\\ 
1 & 0.51 & 0.91 & 0.53 & 0.41 & 0.42 & 0.41 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:279}Sensory data for game 279.}
\end{center}
\end{table}
The hearer guessed the meaning used
by the speaker right away, because both 
share the same perception and both use the most 
salient channel as basis for categorisation. 
\begin{verbatim}
Game 96
  a1 is the speaker. a2 is the hearer. 
  a1 segments the context into 3 objects: 
       rectangle-0 rectangle-1 rectangle-2
  a1 chooses rectangle-1 as the topic 
  a1 categorises the topic as [WIDTH 0.0–0.25]
  a1 creates a new word: `kutaki'
  a1 says: `kutaki'
  a2 does not find a unique referent
  a2 says: `kutaki?'
  a1 points to rectangle-1
  a2 categorises the topic as [WIDTH 0.0–0.25]
  a2 stores `kutaki' as  [WIDTH 0.0–0.25]
\end{verbatim}

@@

Part of the lexicons of five of the ten agents after 2000 games 
(only those words which have positive scores) are shown in the 
\tabref{tab:lex2000}. We see that synonymy does indeed occur. For
example, two words are in the 
running for [\textsc{hpos} 0.5–1.0]: `rutaxese' and `xomupovi', and 
three for [\textsc{vpos} 0.5–1.0]: `wavone', `zaxawe', and 
`dazofo'. Some words are already well established, for
example `numefuli' for [\textsc{gray} 0.5–1.0] (dark).

\begin{table}[h]
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l  }
\lsptoprule
{\itshape Meaning}&{\itshape Word}&{\bfshape  a1}&{\bfshape  a2}&{\bfshape  a3}&{\bfshape  a4}&{\bfshape  a5} \\ \midrule
{}[\textsc{hpos} 0.5–1.0]&rutaxese& &0.1& & &\\ 
 & xomupovi&0.1& &0.2& &0.1\\ 
{}[\textsc{vpos} 0.5–1.0]&wavone& & &0.3& &\\ 
 & zaxawe& & & &0.1& \\ 
 & dazofo&0.1& & &0.2&\\ 
{}[\textsc{width} 0.0–0.5]&buxevo& & & & & \\ 
 & vubupo&1.0&0.6&1.0&0.1&1.0\\ 
{}[\textsc{width} 0.5–1.0]&pawixona& & &0.1& & \\ 
 & rikepule&1.0&0.8&1.0&1.0&1.0\\ 
{}[\textsc{width} 0.5–0.75]&gowinoge& & &0.1& &  \\ 
 & wesurodi& & & &0.2&\\ 
{}[\textsc{width} 0.75–1.0]&besabi& & &0.2& & \\ 
 & lituvi&0.1& & & & \\ 
{}[\textsc{gray} 0.5–0.75]&goxomixe& &0.1& & & \\ 
 & korufo& & & &0.2&\\ 
{}[\textsc{gray} 0.0–0.5]&numefuli&1.0&0.7&1.0&1.0&1.0\\ 
{}[\textsc{gray} 0.0–0.25]&rekemaxi&0.2& & & & \\ 
{}[\textsc{gray} 0.5–1.0]&faluleru&0.6&0.6&1.0&1.0&1.0\\ 
 & nupanu& & & & &0.2 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:lex2000}Lexicon of five agents after 2000 games.}
\end{center}
\end{table}

\clearpage 
The positive feedback loop between use and success
has already dampened some synonyms. For some cases, like [\textsc{gray} 0.5–1.0] the competition has died out 
with one word `faluleru' being the winner. 
For others, like [\textsc{hpos} 0.5–1.0], the competition 
is still going on, although we can guess
that `xomupovi' is probably going to be the winner.

It is instructive to follow the history of the words 
in use for a particular meaning. For 
example, let us look at the words for [\textsc{gray} 0.5–1.0] 
(dark) in the very early phases of lexicon development.
Three different words are quickly created, and 
`faluleru' is the first one that has some success. 
\begin{verbatim}
Game 76
 Speaker a4 creates `notabefe' for [GRAY 0.5–1.0]
 Hearer a5 adopts `notabefe' for [GRAY 0.5–1.0]
Game 79 
 Speaker a2 creates `vivevobo' for [GRAY 0.5–1.0]
 Hearer a7 does not adopt `vivebo' 
     (failed to discriminate)
Game 88 
 Speaker a7 creates `faluleru' for [GRAY 0.5–1.0]
 Hearer a4 adopts `faluleru' for [GRAY 0.5–1.0]
Game 100 
 Speaker a7 uses `faluleru' for [GRAY 0.5–1.0]
 Hearer a4 correctly interprets `faluleru'
\end{verbatim}
The semiotic triangles existing at this point in the population are 
summarised in \figref{triangle6}. 
Agent {\bfshape  a4} is now already in a dilemma because he has created
`notabefe' and picked up `faluleru' from {\bfshape  a7}.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.50\textwidth]{chap6/figs/triangle6.pdf}}
\caption{\label{triangle6}In the case of synonymy, 
there are multiple words for the same meaning and hence the 
same referent.}
\end{figure}

@@

Because there are synonyms, agents must now choose
which word to use. The one with the highest
score should clearly be preferred because based on the
evidence the agent has gathered so far, this
gives the highest chance of success in the game. {\bfshape  a4}
is faced with this kind of choice in 
the next game in the series involving the meaning
{}[\textsc{gray} 0.5–1.0]. {\bfshape  a4} chooses `faluleru' because this 
word has the highest score. It is immediately picked up by 
{\bfshape  a1}: 
\begin{verbatim}
Game 101
  a4 is the speaker. a1 is the hearer. 
  a4 segments the context into 2 objects: 
       rectangle-0 rectangle-1
  a4 chooses rectangle-1 as the topic 
  a4 categorises the topic as [GRAY 0.5–1.0]
  a4 has two words for [GRAY 0.5–1.0]:
       `faluleru' (0.20)
       `notabefe' (0.00)
  a4 says: `faluleru'
  a1 does not know `faluleru'
  a1 says: `faluleru?'
  a4 points to rectangle-1
  a1 categorises the topic as [GRAY 0.5–1.0]
  a1 stores `faluleru' as [GRAY 0.5–1.0]
\end{verbatim}
After this game, three agents `know' the word 
`faluleru' for [\textsc{gray} 0.5–1.0]: {\bfshape  a4}, {\bfshape  a1}, 
and {\bfshape  a7}. When we continue to inspect the
simulation we see that `notabefe' takes a bit of 
a revenge. The next games with the 
meaning [\textsc{gray} 0.5–1.0] all involve the word `notabefe': 
\begin{verbatim}
Game 111
 Speaker a5 uses `notabefe' for [GRAY 0.5–1.0]
 Hearer a4 correctly interprets `notabefe'
Game 113
 Speaker a4 uses `notabefe' for [GRAY 0.5–1.0]
 Hearer a9 adopts `notabefe' for [GRAY 0.5–1.0]
Game 127
 Speaker a9 uses `notabefe' for [GRAY 0.5–1.0]
 Hearer a6 adopts `notabefe' for [GRAY 0.5–1.0]
\end{verbatim}
But then there is another occurrence of `faluleru': 
\begin{verbatim}
Game 135 
 Speaker a1 uses `faluleru' for [GRAY 0.5–1.0]
 Hearer a2 adopts `faluleru' for [GRAY 0.5–1.0]
\end{verbatim}
And now {\bfshape  a6}, which has not been involved 
yet in any interaction concerning the 
meaning [\textsc{gray} 0.5–1.0], further confuses the situation by 
creating a new word, `sopine', which {\bfshape  a5}, playing
the role of hearer, adopts. 
\begin{verbatim}
Game 140
 Speaker a6 creates `sopine' for [GRAY 0.5–1.0]
 Hearer a5 adopts `sopine' for [GRAY 0.5–1.0]
\end{verbatim}
This kind of evolution continues with a struggle between 
`notabefe' and `faluleru'. The agents know both 
words so the games do not fail. But `faluleru', 
just by chance, starts to occur a bit more often
which causes its score to go up a bit more. 
This results in `faluleru' being used even more, 
and, due to lateral inhibition, `notabefe' used 
less. Gradually `faluleru' dominates for 
the whole population. 
After 2000 games, the competitors to `faluleru' have 
all disappeared from the group lexicon.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l }
\lsptoprule
{\itshape Obj} & Hpos & Vpos & Height & Width & Gray & Area \\ \midrule
0 & 0.37 & 0.44 & 0.92 & 0.92 & 0.69 & 0.10\\ 
1 & 0.98 & 0.55 & 0.25 & 0.11 & 0.78 & 0.88\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game3}Sensory data for game 3 after scaling.}
\end{center}
\end{table}
Values for \textsc{vpos} and \textsc{gray} are very close so they are not 
considered as sufficiently salient.
\begin{verbatim}
Game 3
  a1 is the speaker. a2 is the hearer. 
  a1 segments the context into 2 objects: 
       square-0 rectangle-1 
  a1 chooses rectangle-1 as the topic 
  a1 considers as salient AREA WIDTH HEIGHT HPOS 
  a1 categorises the topic as [HEIGHT 0.0–0.5]
  a2 creates a new word: `mibati'
  a1 says: `mibati'
  a2 does not know `mibati'
  a2 says: `mibati?'
  a1 points to rectangle-1
  a1 considers as salient AREA WIDTH HEIGHT HPOS 
  a1 categorises the topic as [HEIGHT 0.0–0.5]
  a1 stores `mibati' as [HEIGHT 0.0–0.5]
\end{verbatim}
{}[\textsc{height} 0.0–0.5] has been chosen by the speaker
because \textsc{height} was one of the salient channels (even 
though clearly not the only salient one) and 
because a successful distinction already existed in 
the ontology. The same 
distinction was chosen by chance by the hearer, but he 
could just as well have chosen a distinction based on 
the \textsc{area}, \textsc{width} or \textsc{hpos}.

In the next game of the simulation series, `mibati'
is used again, now by {\bfshape  a2} as speaker. It is 
based on a scene with a triangle and a rectangle.
The data for game 4 are shown in \tabref{tab:mibati} (after scaling).

>>>>

The speaker first plays a Discrimination Game traversing the 
semantic side of the square going from the referent 
{\itshape rectangle-1} to a possible meaning [\textsc{gray} 0.5–1.0]. 
He then plays a Naming Game traversing the lexical side of 
of the square to find the word \smplenquote{pokuneso} for this chosen meaning. 
The hearer traverses the lexical side of the triangle in 
the other direction to interpret the word \smplenquote{pokuneso} as
{}[\textsc{gray} 0.5–1.0], and then identifies the referent by 
filtering the objects in the context with this meaning. 
Only rectangle-1 remains, so the game succeeds. 
The whole game is reported by the commentator as follows: 
\begin{verbatim}
Game 500
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 3 objects: 
       rectangle-0 rectangle-1 rectangle-2
  a2 chooses rectangle-1 as the topic 
  a2 categorises the topic as [GRAY 0.5–1.0]
  a2 says: pokuneso
  a1 interprets pokuneso as [GRAY 0.5–1.0]
  a1 points to rectangle-1
  a2 signals OK 
\end{verbatim}
The game is perfectly successful because both 
agents associate the word \smplenquote{okuneso} with 
{}[\textsc{gray} 0.5–1.0] (dark) and they perceive the scene 
in the same way.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l }
\lsptoprule
{\itshape Meaning}&{\itshape Word}&{\itshape Translation} & {\bfshape  a1}&{\bfshape  a2} \\ \midrule
{}[\textsc{hpos} 0.0–0.5] & vapola&left&-&0.1\\ 
{}[\textsc{hpos} 0.5–1.0]& gonapa&right &0.1&-\\ 
{}[\textsc{height} 0.0–0.5]&suwaxugo&short &0.6&0.8\\ 
{}[\textsc{height} 0.5–1.0]& kusone&tall &0.4&0.5\\ 
{}[\textsc{width} 0.0–0.5]&bepupepa&narrow &0.1&0.1\\ 
{}[\textsc{width} 0.0–0.25]&kutaki&very narrow &-&0.1\\ 
{}[\textsc{width} 0.5–1.0]& zikorika&wide &0.0&0.3\\ 
{}[\textsc{gray} 0.0–0.5]& fesasado&light &0.5&0.7\\ 
{}[\textsc{gray} 0.5–1.0]& pokuneso&dark &0.8&0.9\\ 
{}[\textsc{area} 0.5–1.0]& mafanoda&large &0.1&0.1\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:lex100}. Complete lexicon of {\bfshape  a1} and {\bfshape  a2} after 100 games.}
\end{center}
\end{table}
We see that, at this point, the agents have lexicalised 
only the most general distinctions, such as \smplenquote{dark} (\emph{pokuneso}) 
versus \smplenquote{light} (\emph{fesasado}) or \smplenquote{short} (\emph{suwaxugo}) versus \smplenquote{tall} 
(\emph{kusone}). Words for the grayscale and height dimensions
have the strongest scores, although this is purely accidental. 
When we would start another simulation from scratch,
we would end up with different words and perhaps 
other distinctions would be more successful.

@@

We see that words for  basic distinctions have further established
themselves. Words for \smplenquote{short} (\emph{suwaxugo}) and \smplenquote{tall} (\emph{kusone}), or \smplenquote{light} (\emph{fesasado}) and \smplenquote{dark} (\emph{pokuneso}) now have scores of 1.0. Words for more refined categories, like \smplenquote{very short} (\emph{tawube}) or \smplenquote{very narrow} (\emph{kutaki}), are beginning to establish themselves.

@@

Note that this simulation is very different from the 
ones shown in the previous chapter. The agents now get 
only feedback through overt selection of the referent. 
The hearer points to the identified referent and the speaker 
decides on the outcome of the game based on this 
non-verbal information, but speaker and hearer do not 
know whether they have used the same meaning or not. 
Very often there are alternative ways
to conceptualise reality, so even if agents would have
completely shared ontologies, there is still the possibility 
of guessing the wrong meaning. I have called this
the gavagai-problem, inspired by the philosopher
Quine, who tells the story of the anthropologist 
puzzled by the word \smplenquote{gavagai} uttered by a native 
in an undecoded language. Does \smplenquote{gavagai} mean rabit, animal 
scurrying by, the direction in which I will go 
now, or white furry object? 
The child who is acquiring a lexicon has
exactly the same problem. It explains why 
overextensions or underextensions
are seen in a child's first words. For example, 
the word for orange is applied to any small circular round 
object, including a ball, or a doorknob.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l }
\lsptoprule
{\itshape Obj} & Hpos & Vpos & Height & Width & Gray & Area \\ \midrule
0 & 0.51 & 0.98 & 0.90 & 0.47 & 0.79 & 0.60\\ 
1 & 0.75 & 0.68 & 0.26 & 0.09 & 0.14 & 0.20\\ 
2 & 0.76 & 0.54 & 0.56 & 0.26 & 0.94 & 0.36\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game77}Sensory data for game 77 after sensor-scaling.}
\end{center}
\end{table}
The speaker is again {\bfshape  a2}. \textsc{height} is the most salient channel. 
{\bfshape  a2} has a word for [\textsc{height} 0.75–1.0] (very tall), namely
\smplenquote{bowaluro}, and uses it. The hearer {\bfshape  a1} 
does not know the 
word, conceptualises the scene, and arrives at the 
same category, so the hearer stores the new word
with the same meaning as the speaker. 
\begin{verbatim}
Game 77
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 3 objects: 
       rectangle-0 rectangle-1 rectangle-2
  a2 chooses rectangle-1 as the topic 
  a2 categorises the topic as [HEIGHT 0.75–1.0]
  a2 says: \smplenquote{bowaluro}
  a1 does not know \smplenquote{bowaluro}
  a1 says: \smplenquote{bowaluro?}
  a2 points to rectangle-1
  a1 categorises the topic as [HEIGHT 0.75–1.0]
  a1 stores \smplenquote{bowaluro} as [HEIGHT 0.75–1.0]
\end{verbatim}
The reason why {\bfshape  a1} has guessed the right 
meaning of \smplenquote{bowaluro} is because both agents use
only the most salient channel and they both share
the same perception of reality. We will soon see 
that if these constraints are not valid, agents 
are not always so lucky and hence multiple meanings
start to circulate for the same word.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l }
\lsptoprule
{\itshape Obj}&Hpos&Vpos&Height&Width&Gray&Area \\ \midrule
0 &0.51 & 0.71 & 0.64 & 0.61 & 0.80 & 0.56\\ 
1 & 0.51 & 0.91 & 0.53 & 0.41 & 0.42 & 0.41 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:279}Sensory data for game 279.}
\end{center}
\end{table}
The hearer guessed the meaning used
by the speaker right away, because both 
share the same perception and both use the most 
salient channel as basis for categorisation. 
\begin{verbatim}
Game 96
  a1 is the speaker. a2 is the hearer. 
  a1 segments the context into 3 objects: 
       rectangle-0 rectangle-1 rectangle-2
  a1 chooses rectangle-1 as the topic 
  a1 categorises the topic as [WIDTH 0.0–0.25]
  a1 creates a new word: \smplenquote{kutaki}
  a1 says: \smplenquote{kutaki}
  a2 does not find a unique referent
  a2 says: \smplenquote{kutaki?}
  a1 points to rectangle-1
  a2 categorises the topic as [WIDTH 0.0–0.25]
  a2 stores \smplenquote{kutaki} as  [WIDTH 0.0–0.25]
\end{verbatim}

@@

Part of the lexicons of five of the ten agents after 2000 games 
(only those words which have positive scores) are shown in the 
\tabref{tab:lex2000}. We see that synonymy does indeed occur. For
example, two words are in the 
running for [\textsc{hpos} 0.5–1.0]: \smplenquote{rutaxese} and \smplenquote{xomupovi}, and 
three for [\textsc{vpos} 0.5–1.0]: \smplenquote{wavone}, \smplenquote{zaxawe}, and 
\smplenquote{dazofo}. Some words are already well established, for
example \smplenquote{numefuli} for [\textsc{gray} 0.5–1.0] (dark).

\begin{table}[p]
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l  }
\lsptoprule
{\itshape Meaning}&{\itshape Word}&{\bfshape  a1}&{\bfshape  a2}&{\bfshape  a3}&{\bfshape  a4}&{\bfshape  a5} \\ \midrule
{}[\textsc{hpos} 0.5–1.0]&rutaxese& &0.1& & &\\ 
 & xomupovi&0.1& &0.2& &0.1\\ 
{}[\textsc{vpos} 0.5–1.0]&wavone& & &0.3& &\\ 
 & zaxawe& & & &0.1& \\ 
 & dazofo&0.1& & &0.2&\\ 
{}[\textsc{width} 0.0–0.5]&buxevo& & & & & \\ 
 & vubupo&1.0&0.6&1.0&0.1&1.0\\ 
{}[\textsc{width} 0.5–1.0]&pawixona& & &0.1& & \\ 
 & rikepule&1.0&0.8&1.0&1.0&1.0\\ 
{}[\textsc{width} 0.5–0.75]&gowinoge& & &0.1& &  \\ 
 & wesurodi& & & &0.2&\\ 
{}[\textsc{width} 0.75–1.0]&besabi& & &0.2& & \\ 
 & lituvi&0.1& & & & \\ 
{}[\textsc{gray} 0.5–0.75]&goxomixe& &0.1& & & \\ 
 & korufo& & & &0.2&\\ 
{}[\textsc{gray} 0.0–0.5]&numefuli&1.0&0.7&1.0&1.0&1.0\\ 
{}[\textsc{gray} 0.0–0.25]&rekemaxi&0.2& & & & \\ 
{}[\textsc{gray} 0.5–1.0]&faluleru&0.6&0.6&1.0&1.0&1.0\\ 
 & nupanu& & & & &0.2 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:lex2000}Lexicon of five agents after 2000 games.}
\end{center}
\end{table}

\begin{figure}[p]
  \centerline{\includegraphics[width=.50\textwidth]{chap6/figs/triangle6.pdf}}
\caption{\label{triangle6}In the case of synonymy, 
there are multiple words for the same meaning and hence the 
same referent.}
\end{figure}

\clearpage 
The positive feedback loop between use and success
has already dampened some synonyms. For some cases, like [\textsc{gray} 0.5–1.0] the competition has died out 
with one word \smplenquote{faluleru} being the winner. 
For others, like [\textsc{hpos} 0.5–1.0], the competition 
is still going on, although we can guess
that \smplenquote{xomupovi} is probably going to be the winner.

It is instructive to follow the history of the words 
in use for a particular meaning. For 
example, let us look at the words for [\textsc{gray} 0.5–1.0] 
(dark) in the very early phases of lexicon development.
Three different words are quickly created, and 
\smplenquote{faluleru} is the first one that has some success. 
\begin{verbatim}
Game 76
 Speaker a4 creates \smplenquote{notabefe} for [GRAY 0.5–1.0]
 Hearer a5 adopts \smplenquote{notabefe} for [GRAY 0.5–1.0]
Game 79 
 Speaker a2 creates \smplenquote{vivevobo} for [GRAY 0.5–1.0]
 Hearer a7 does not adopt \smplenquote{vivebo} 
     (failed to discriminate)
Game 88 
 Speaker a7 creates \smplenquote{faluleru} for [GRAY 0.5–1.0]
 Hearer a4 adopts \smplenquote{faluleru} for [GRAY 0.5–1.0]
Game 100 
 Speaker a7 uses \smplenquote{faluleru} for [GRAY 0.5–1.0]
 Hearer a4 correctly interprets \smplenquote{faluleru}
\end{verbatim}
The semiotic triangles existing at this point in the population are 
summarised in \figref{triangle6}. 
Agent {\bfshape  a4} is now already in a dilemma because he has created
\smplenquote{notabefe} and picked up \smplenquote{faluleru} from {\bfshape  a7}.

@@

Because there are synonyms, agents must now choose
which word to use. The one with the highest
score should clearly be preferred because based on the
evidence the agent has gathered so far, this
gives the highest chance of success in the game. {\bfshape  a4}
is faced with this kind of choice in 
the next game in the series involving the meaning
{}[\textsc{gray} 0.5–1.0]. {\bfshape  a4} chooses \smplenquote{faluleru} because this 
word has the highest score. It is immediately picked up by 
{\bfshape  a1}: 
\begin{verbatim}
Game 101
  a4 is the speaker. a1 is the hearer. 
  a4 segments the context into 2 objects: 
       rectangle-0 rectangle-1
  a4 chooses rectangle-1 as the topic 
  a4 categorises the topic as [GRAY 0.5–1.0]
  a4 has two words for [GRAY 0.5–1.0]:
       \smplenquote{faluleru} (0.20)
       \smplenquote{notabefe} (0.00)
  a4 says: \smplenquote{faluleru}
  a1 does not know \smplenquote{faluleru}
  a1 says: \smplenquote{faluleru?}
  a4 points to rectangle-1
  a1 categorises the topic as [GRAY 0.5–1.0]
  a1 stores \smplenquote{faluleru} as [GRAY 0.5–1.0]
\end{verbatim}
After this game, three agents \smplenquote{know} the word 
\smplenquote{faluleru} for [\textsc{gray} 0.5–1.0]: {\bfshape  a4}, {\bfshape  a1}, 
and {\bfshape  a7}. When we continue to inspect the
simulation we see that \smplenquote{notabefe} takes a bit of 
a revenge. The next games with the 
meaning [\textsc{gray} 0.5–1.0] all involve the word \smplenquote{notabefe}:

\begin{verbatim}
Game 111
 Speaker a5 uses \smplenquote{notabefe} for [GRAY 0.5–1.0]
 Hearer a4 correctly interprets \smplenquote{notabefe}
Game 113
 Speaker a4 uses \smplenquote{notabefe} for [GRAY 0.5–1.0]
 Hearer a9 adopts \smplenquote{notabefe} for [GRAY 0.5–1.0]
Game 127
 Speaker a9 uses \smplenquote{notabefe} for [GRAY 0.5–1.0]
 Hearer a6 adopts \smplenquote{notabefe} for [GRAY 0.5–1.0]
\end{verbatim}
But then there is another occurrence of \smplenquote{faluleru}: 
\begin{verbatim}
Game 135 
 Speaker a1 uses \smplenquote{faluleru} for [GRAY 0.5–1.0]
 Hearer a2 adopts \smplenquote{faluleru} for [GRAY 0.5–1.0]
\end{verbatim}
And now {\bfshape  a6}, which has not been involved 
yet in any interaction concerning the 
meaning [\textsc{gray} 0.5–1.0], further confuses the situation by 
creating a new word, \smplenquote{sopine}, which {\bfshape  a5}, playing
the role of hearer, adopts. 
\begin{verbatim}
Game 140
 Speaker a6 creates \smplenquote{sopine} for [GRAY 0.5–1.0]
 Hearer a5 adopts \smplenquote{sopine} for [GRAY 0.5–1.0]
\end{verbatim}
This kind of evolution continues with a struggle between 
\smplenquote{notabefe} and \smplenquote{faluleru}. The agents know both 
words so the games do not fail. But \smplenquote{faluleru}, 
just by chance, starts to occur a bit more often
which causes its score to go up a bit more. 
This results in \smplenquote{faluleru} being used even more, 
and, due to lateral inhibition, \smplenquote{notabefe} used 
less. Gradually \smplenquote{faluleru} dominates for 
the whole population. 
After 2000 games, the competitors to \smplenquote{faluleru} have 
all disappeared from the group lexicon.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l  l  l }
\lsptoprule
{\itshape Obj} & Hpos & Vpos & Height & Width & Gray & Area \\ \midrule
0 & 0.37 & 0.44 & 0.92 & 0.92 & 0.69 & 0.10\\ 
1 & 0.98 & 0.55 & 0.25 & 0.11 & 0.78 & 0.88\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game3}Sensory data for game 3 after scaling.}
\end{center}
\end{table}
Values for \textsc{vpos} and \textsc{gray} are very close so they are not 
considered as sufficiently salient.
\begin{verbatim}
Game 3
  a1 is the speaker. a2 is the hearer. 
  a1 segments the context into 2 objects: 
       square-0 rectangle-1 
  a1 chooses rectangle-1 as the topic 
  a1 considers as salient AREA WIDTH HEIGHT HPOS 
  a1 categorises the topic as [HEIGHT 0.0–0.5]
  a2 creates a new word: \smplenquote{mibati}
  a1 says: \smplenquote{mibati}
  a2 does not know \smplenquote{mibati}
  a2 says: \smplenquote{mibati?}
  a1 points to rectangle-1
  a1 considers as salient AREA WIDTH HEIGHT HPOS 
  a1 categorises the topic as [HEIGHT 0.0–0.5]
  a1 stores \smplenquote{mibati} as [HEIGHT 0.0–0.5]
\end{verbatim}
{}[\textsc{height} 0.0–0.5] has been chosen by the speaker
because \textsc{height} was one of the salient channels (even 
though clearly not the only salient one) and 
because a successful distinction already existed in 
the ontology. The same 
distinction was chosen by chance by the hearer, but he 
could just as well have chosen a distinction based on 
the \textsc{area}, \textsc{width} or \textsc{hpos}.

In the next game of the simulation series, \smplenquote{mibati}
is used again, now by {\bfshape  a2} as speaker. It is 
based on a scene with a triangle and a rectangle.
The data for game 4 are shown in \tabref{tab:mibati} (after scaling).
